Experience Report: Haskell in Computational Biology

### 2012-02-27 mrfy_experience_report.tex
a few code example improvements. Added transform of Viterbi relations in negative log space.

\conferenceinfo{ICFP '12}{date, City.} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

@@

\title{Experience Report: Haskell in Computational Biology}
\subtitle{Subtitle Text, if any}

@@

\begin{figure}
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}}
\caption{The directed graph representing the states of a MRFy Markov random field finite state
automaton}\label{mrf}
\end{figure}

@@

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations that seek to maximize the probability of an observed sequence (in 
this case, of amino acid ``letters'') being emitted by a particular 
finite-state automaton (the hidden Markov model). The particular states of this 
finite-state automaton map to the recurrence relations, and in this instance 
derive from what is known in the computational biology community as a 
\textit{profile} hidden Markov model. The Viterbi algorithm in this instance 
recurses on ``match,'' ``insertion,'' and ``deletion'' states. To be efficient, 
the Viterbi algorithm relies on dynamic programming, or memoization, and the 
resulting asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence 
of distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.

@@

@@

 - SMURF bad
 - HMMER good

@@

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

@@

Acknowledgments, if needed.

>>>>

\conferenceinfo{ICFP '12}{September 9-15, Copenhagen.} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

@@

\title{Experience Report: Haskell in Computational Biology}
% \subtitle{Subtitle Text, if any}

@@

\begin{figure} 
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\caption{The directed graph representing the states of a MRFy Markov random 
field finite state automaton. Shaded nodes represent beta-strand positions, 
while the dashed edges represent the long-range dependencies due to hydrogen 
bonding.}\label{mrf} \end{figure}

@@

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model). 
The particular states of this finite-state automaton map to the recurrence 
relations, and in this instance derive from what is known in the computational 
biology community as a \textit{profile} hidden Markov model. The Viterbi 
algorithm in this instance recurses on ``match,'' ``insertion,'' and 
``deletion'' states. To be efficient, the Viterbi algorithm relies on dynamic 
programming, or memoization, and the resulting asymptotic complexity is 
$O(|M|\times|N|)$, when $M$ is the sequence of distinct states (nodes in the 
automaton's graph) and $N$ is the sequence of letters in the protein sequence. 
In our particular problem domain, both $M$ and $N$ routinely contain several 
hundred to a few thousand elements.

@@

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}I_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

@@

Haskell makes it difficult to write code in haste, because Haskell militates towards careful type-checking and clearly thought-out function contracts. Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the programmer who attempts to write code quickly with the goal of later making it work correctly. We have found MRFy to be easier to enhance and maintain than our existing C++ code bases.

@@

% \appendix
% \section{Appendix Title}
% 
% This is the text of the appendix, if you need one.

@@

We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford Larsen, and Nathan Ricci.
